Addressing climate change and development pressures in an urban estuary through habitat restoration planning

### 2022-08-31 manu-draft.qmd
adding area estimates to text for opp area summary

After binning, the native and restorable lands were overlaid with additional layers to identify 1) reservation native and reservation restorable areas, and 2) existing and proposed native and restorable areas. Collectively, each of these unique products are considered the opportunity areas in the Tampa Bay watershed. The reservation native and reservation restorable areas are native and restorable habitats, respectively, that occur in the 5-foot contour or coastal stratum. This area extends from the local Mean Lower Low Water elevation to elevation 5 feet above Mean Sea Level and is likely to be affected by frequent tidal flooding or inundation [@tbep0519]. Native habitats in this stratum were identified as those to be reserved, whereas restorable habitats were identified as those where tidal wetlands could be restored.

@@

The geospatial analysis that produced a map of the remaining opportunity areas provided an areal estimate of where practitioners could target future restoration projects (@fig-oppmap). Native lands categorized as existing conservation native, proposed conservation native, or reservation native totaled.  Similarly, restorable lands categorized as existing conservation restorable, proposed conservation restorable, or reservation restorable totalled.  Understandably, most of the native and restorable lands occurred in undeveloped areas of the watershed, particularly in the north and southwest. Existing conservation lands (existing conservations native, existing conservation restorable) totalled and proposed conservation lands (proposed conservation native, proposed conservation restorable).  Areas in the coastal stratum (the MLLW to 5-foot contour) included of reservation native habitats and ha of reservation restorable habitats.

Combining the restorable lands in @fig-oppmap with the coastal stratum, soils, and salinity data provided a spatial summary of the restoration potential grouped by habitat type (@fig-restmap).  The restortation potential
In 2017, 1,555 km^2^ of land (26.5%) in the Tampa Bay watershed above the MLLW line was classified as natural and 2,144 km^2^ (37%) was considered restorable. Developed areas in the Tampa Bay watershed encompassed 2,172 km^2^ (36.5%) of the 5,872 km^2^ watershed area. Between 2014 and 2017, the developed footprint increased by seven percent (7%, Fig 2).

The Tampa Bay watershed included a total of 1,260 km^2^ of existing conservation lands, either publicly owned or in conservation easements. However, excluding subtidal areas owned by the State of Florida results in a total of 816 km^2^ of conservation lands, about 13.9% of the watershed area occurring above the MLLW line. (Fig 3). Proposed conservation lands in the Tampa Bay watershed total 1,254 km^2^. The mapped proposed conservation lands generally link eastward and provide wildlife connectivity to the larger-scale Florida Wildlife Corridor (Fig 3).

Reservation lands, from the MLLW line landward to elevation 5-feet (NAVD 88), currently (2017) include 4,622 ha proposed for conservation in the Tampa Bay watershed, with 3,303 ha of native habitats and 1,319 ha of restorable habitats (Fig 1).

@fig-restmap

>>>>

After binning, the native and restorable lands were overlaid with additional layers to identify 1) reservation native and reservation restorable areas, and 2) existing and proposed native and restorable areas. Collectively, each of these unique products are considered the opportunity areas in the Tampa Bay watershed. The reservation native and reservation restorable areas are native and restorable habitats, respectively, that occur in the 5-foot contour or coastal stratum. This area extends from the local Mean Lower Low Water (MLLW) elevation to elevation 5 feet above Mean Sea Level and is likely to be affected by frequent tidal flooding or inundation [@tbep0519]. Native habitats in this stratum were identified as those to be reserved, whereas restorable habitats were identified as those where tidal wetlands could be restored.

@@

The map of the remaining opportunity areas provided an areal estimate of where practitioners could target future restoration projects (@fig-oppmap). Native lands (existing conservation native, proposed conservation native, reservation native) totaled `r areas$native$areaha` (`r areas$native$areaper` of the watershed above MLLW).  Similarly, restorable lands (existing conservation restorable, proposed conservation restorable, reservation restorable) totalled `r areas$restorable$areaha` (`r areas$restorable$areaper` of watershed).  Understandably, most of the native and restorable lands occurred in undeveloped areas of the watershed, particularly in the north and southwest. Existing conservation lands (existing conservation native, existing conservation restorable) totalled `r areas$existing$areaha` (`r areas$existing$areaper` of the watershed) and proposed conservation lands (proposed conservation native, proposed conservation restorable) totalled `r areas$proposed$areaha` (`r areas$proposed$areaper` of the watershed).  Reservation areas in the coastal stratum (reservation native, reservation restorable) totalled `r areas$reservation$areaha` (`r areas$reservation$areaper` of the watershed).

Combining the restorable lands in @fig-oppmap with the coastal stratum, soils, and salinity data provided a spatial summary of the restoration potential grouped by habitat type (@fig-restmap).
